---
title: "Finding the story in salaries data"
author: "Liz Lucas @ IRE, Ryan Thornburg"
output: html_document
---

Now you'll put into practice the functions you've learned so far to interrogate some salary data from Bloomington, Indiana, that came from a records request. We have cleaned up the data a little for the purposes of this class, but left it in spreadsheet format, so shortly you'll
learn how to import data from an Excel file (either .xls or .xlsx).

First, open up `salaries.xlsx` in Excel by double-clicking on the file in Finder. Note that it has two tabs: one with the data, an another with notes on the source. This is best practice for keeping track of when and where you received data. But you only want to import the first tab into R for analysis.

### Part One: Installing Packages

When you first install R on your computer it comes with a wide variety of functions that are collectively called "base R."

Base R works just fine, and you'll often find that the base functions do what you want more clearly or efficiently than anything else. But to do the really cool stuff, you'll need to download additional packages. Each package comes with new verbs that extend base R and often provide shortcuts or additional actions.

You can think of installing packages as adding new verbs to your vocabulary. Think of base R as giving you the verb "turn on." If you were to install a hypothetical "southernisms" package, you might get a verb called "cut on." Different verbs. Same action.

In other cases, packages give you new verbs that do a bunch of things at once. For example, imagine a theoretical package called "PBJ" that has a function called `make_sandwich()` . In base R you might have to `get_bread()` and then `spread_pb()` and then `spread_jelly()`. It's much nicer to just `make_sandwich()`.

The first package that we'll install is called tidyverse. Its really a package of packages that are used all ... the ... time. You can read more about the tidyverse and its core packages at <https://www.tidyverse.org/packages/>

Installing a package in R means downloading the code from a remote location on to your local computer. Most often, you are downloading packages from something called "CRAN," which is "a network of ftp and web servers around the world that store identical, up-to-date, versions of code and documentation for R." You can also install packages from GitHub, but that's a story for a different class.

To install a package that's on CRAN, we use the `install.packages()` function from base R. The argument is the name of the package as a string, contained inside quote marks.

I've commented out this line because IRE has already downloaded the packages on to the conference machines. 

```{r}
#install.packages("tidyverse")
```

When you run that, you will see some action down in your console. You can also verify that the package has been installed by going over to the "Packages" tab in R Studio.

**You only need to install packages once per computer.**

Even though you only need to install a package once, you need to load it every time you start a new R session. To do that, we use the `library()` function.

```{r}
library(tidyverse)
```

To verify that the package has been correctly loaded into your library for the current R session, hop on over to the Packages tab in R Studio and make sure there's a check mark next to the name of the package you intended to load.

If you're looking for a short-cut on installing packages, you can use the "Install" button in the "Packages" tab in R Studio. This is probably OK to use because you only need to install packages once. It's not code you need to run repeatedly.

There's also a shortcut for loading packages into your library each session. You can just click or un-click the check box next to the package name in R Studio. But this isn't a good habit to get into. You typically want a code chunk at the top of your Rmd file that loads all the packages you need to run the rest of the work. That way you can just run the chunk each time. More importantly, it makes clear to other folks which packages they will need if they want to run your script on their computer.

Finally, there's a useful shortcut for installing and loading packages. It's in the `pacman` package, and it's called `p_load()` .

I like it for two reasons. It's an easy way to load multiple packages at once, and it installs any packages you haven't already downloaded on to your computer.

```{r}
#install.packages("pacman")
```

```{r}
pacman::p_load(tidyverse, readxl)
```

### Part Two: Loading Data

#### Loading CSV files
Base R has a function called `read.csv()` that is used to -- you guessed it -- read a csv file into R. But we're going to use a very similarly named -- but better -- function from the tidyverse called `read_csv()`. While only one character differentiates these two functions, and while they both have the same end goal, these two functions work very differently. If you google a solution that uses read.csv, it may not work if you're using read_csv.

Let's first use base R.
```{r}
teammentions_base_R <- read.csv("../data/teammentions.csv",
                              header = TRUE, 
                              sep=",",
                              quote = "\""
                              )
```

Here's the tidyverse way of doing the same thing.

```{r}
teammentions <- read_csv("../data/teammentions.csv", 
    col_types = cols(
      DATE = col_datetime(
               format = "%m/%d/%Y %H:%M")
      )
    )
```
One advantage of using the tidyverse method is that you can ensure that date columns are correctly read in as dates rather than character strings. So why the error? Three of the rows didn't have hours and minutes, which the formatting argument was expecting.

#### Loading Excel Spreadsheets
There are many functions available in `readxl`, the one you'll use now is read_excel(). This function has an optional argument called "sheet," which allows you to specify, numerically, which sheet or sheets you want to import. We want the first one:

```{r}
read_excel("../data/salaries.xlsx", sheet=1)
```

**Remember!** The results of any function - including read_excel() - either print to the console or save to a variable. If you want to refer to this data table later and pipe it into functions, you need to save it to a variable. Call it "salaries":

```{r}
salaries <- read_excel("../data/salaries.xlsx", sheet=1)
```

But remember what we said about naming variables and objects? Those last four are going to make your life miserable. 

We can use a function called `clean_names()` from the very useful `janitor` package to make a nice fix.
```{r}
#install.packages("janitor")
salaries_clean_names <- janitor::clean_names(salaries)
```

You can fix that last column name with the `rename()` function from the tidyverse.
```{r}
salaries_clean_names <- salaries_clean_names %>%
  rename("salary_2021" = "x2021_salary")
```

And if you like it you can go ahead and overwrite the original data frame and delete the old one by using the broom icon in R Studio's Environment tab.
```{r}
salaries <- salaries_clean_names
```


#### Two cheats
Loading data has two short cuts that are similar to those we used for installing and loading packages.

First, we can use the "Import Dataset" button in the "Environment" tab of R Studio. This is a nice way to get started learning some of the nuances of loading data because it visually walks you thorugh options and then generates code you can cut and paste into your Rmd file. However, you always want to make sure that the code you need to load the data is included in your Rmd. Omitting it can cause all sorts of problems for other users of your code as well as future you.

There is also a coding short cut that uses the `import()` function from the `rio` package.The advantage of this function is that it's the same whether you're pulling in a csv file, an Excel spreadsheet or 34 other data types.

In this example, we import only the first of the .txt files.
```{r}
#install.packages("rio")
baby_names_2010 <- rio::import("https://www.ssa.gov/oact/babynames/names.zip", which = "yob2010.txt")
```

Or we can import all of them and "bind" the rows from each file into a single data frame:
```{r}
baby_names_all <- rio::import_list("https://www.ssa.gov/oact/babynames/names.zip", rbind = TRUE)
```

Sometimes you need to add column names. This is a case in which base R may be more efficient than the tidyverse method.
```{r}
#tidyverse method
baby_names_2010 <- baby_names_2010 %>%
  rename("name" ="V1",
         "sex" = "V2",
         "babies" = "V3")
```

```{r}
#base R method
colnames(baby_names_all) <- c("name", "sex", "babies", "year")
```


### Part Three: Taking Your Data Out to Coffee

Later, you're going to learn how to really interview your data. But just like human sources, it's good just to get to know them first before you need to interview them for a story.

**Before you interview your data, you want to take it out for coffee.**

When you take your data out for coffee, you're trying to assess two things:

1.  What does your data know and what does it not know?

2.  What might it try to mislead you about?

Take a look at the salaries data: click on the word "salaries" in your Environment (upper right). Take a minute or two to look at the data:
- What is one row of data? (One employee) 
- What columns of information do you have?

We can also use functions to get to know our data.

-   `str()` – tells you the length and width of the data table as well as the data type for each column

-   `summary()` – in addition to telling you the data type for each column, for numerical columns it will provide some summary statistics and tell you how many `NA` values you have

-   `Hmisc::describe()` - adds percentiles for numerical data, and also tells provides the number of unique values for text columns, and also provides summary statistics for date columns

-   `skimr::skim()` - for numerical values, provides a histogram so you can roughly see if the data skews one way or the other

```{r}
str(salaries)
```

```{r}
summary(salaries)
```

```{r}
#install.packages("Hmisc")
Hmisc::describe(salaries)
```

```{r}
#install.packages("skimr")
skim <- skimr::skim(salaries)
View(skim)
```

Note that there are NAs in the overtime_oncall, hourly_rate, and salary_2021 columns. NAs are *NULL* values, not blanks.

#### Your turn

Start with some basic questions:

-   How many employees in our data?

-   What was the highest total compensation?

-   How many distinct job titles are we dealing with?

-   How much does an employee have to make to be in the top 10 percent of salaries?

-   What percentage of the city's staff is salaried?

-   What's the median number of weekly hours worked?

### Part Four: Summarising Columns

That's right. I spelled summarise with an "s" instead of a "z". And so should you. The person who developed the tidyverse is from New Zeland. And while he kindly allows both `summarise()` and `summarize()` to work, you will find that the American spelling of the word conflicts with other functions in other packages. (The Hmisc package has a summarize() function.)

```{r}
#Hey R, start with the salaries data frame... 
salaries %>% # ... and then ... 
  #... add up all the values in the total_comp column and print that value under a column header called citywide_payroll.
  summarise(citywide_payroll = sum(total_comp))
```

#### Your turn

-   What's the median salary?

```{r}
____________
  _________(median_salary ____________________)
  
# You should get $37,750.26.  
```

-   How much did the city spend on overtime/oncall pay?

```{r}


```

Here's where NAs (NULLs) will trip you up. If you sum a column with NAs in it, R will return an NA. So you need to exclude the NAs in your summing. Thankfully there is an EASY way to do this; the sum() function will take an additional argument: `na.rm=T`, which means remove NAs.

Adding it looks like this:

```{r}
salaries %>% 
  summarise(total_payroll = sum(overtime_oncall, na.rm=T))
```

That's why it's important to take note of NAs in your data! Anytime you want to sum a column with NAs, you need to include this argument in the aggregate function: `na.rm=T`

-   What's the average and median salary for 2021? Hourly rate? (Note: both of these have NAs, so code accordingly)

```{r}

```
